---
title: "STAT101C Final Regression Projet"
output:
  pdf_document: default
  html_document: default
date: "2025-12-06"
authors: Bebel Yen, Victoria Yee, Kaitlyn Wu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Load Libraries

```{r}
library(tidyverse)
library(lubridate)
library(randomForest)
library(glmnet)
library(xgboost)
```

# 1. Load Data 

```{r}
survey <- read_csv("survey_train_test.csv")
purch <- read_csv("amazon-purchases.csv")

purch <- purch %>%
  rename(Survey.ResponseID = `Survey ResponseID`)
```

# 2. Data Preprocessing

## Date Processing and Line Total

```{r}
purch <- purch %>%
  mutate(
    `Order Date` = mdy(`Order Date`),
    line_total = `Purchase Price Per Unit` * Quantity
  )

# head(purch)
```

## Income Distribution Across Amazon Usage Frequency
```{r}
chart_data <- survey %>%
  filter(!is.na(Q.demos.income), !is.na(Q.amazon.use.how.oft)) %>%
  count(Q.amazon.use.how.oft, Q.demos.income) %>%
  group_by(Q.amazon.use.how.oft) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(chart_data, aes(x = Q.amazon.use.how.oft, y = prop, fill = Q.demos.income)) +
  geom_col(width = 0.7) + 
  labs(title = "Income Distribution Across Amazon Usage Frequency", 
       y = "Amazon Usage Frequency", x = "Proportion") + 
  theme_minimal()
```

## Distribution of Household Size
```{r, fig.height = 3}
ggplot(survey, aes(x = Q.amazon.use.hh.size.num)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "white") +
  labs(
    title = "Distribution of Household Size",
    x = "Household Size",
    y = "Count"
  ) +
  theme_minimal()
```

# 3. Feature Engineering (Bebel Yen)

## Basic Aggregates Per User

```{r}
agg_num <- purch %>%
  group_by(Survey.ResponseID) %>%
  summarise(
    n_orders = n(),
    total_quantity = sum(Quantity, na.rm = TRUE),
    total_spend = sum(line_total, na.rm = TRUE),
    mean_price = mean(`Purchase Price Per Unit`, na.rm = TRUE),
    max_price = max(`Purchase Price Per Unit`, na.rm = TRUE),
    min_price = min(`Purchase Price Per Unit`, na.rm = TRUE),
    first_order_date = min(`Order Date`, na.rm = TRUE),
    last_order_date = max(`Order Date`, na.rm = TRUE),
    n_unique_states = n_distinct(`Shipping Address State`),
    n_unique_titles = n_distinct(Title),
    n_unique_asins = n_distinct(`ASIN/ISBN (Product Code)`),
    n_unique_cats = n_distinct(Category)
  ) %>%
  ungroup()

# head(agg_num)
```

## Duration and Recency Features

```{r}
max_date <- max(purch$`Order Date`, na.rm = TRUE)

agg_num <- agg_num %>%
  mutate(
    days_active = as.numeric(difftime(last_order_date, first_order_date, units = "days")),
    days_active = replace_na(days_active, 0),
    days_since_last_order = as.numeric(difftime(max_date, last_order_date, units = "days"))
  ) %>%
  select(-first_order_date, -last_order_date)

# head(agg_num)
```

```{r}
par(mfrow = c(1, 2))
hist(agg_num$total_spend, col = "blue", main = "Total Spend")
hist(agg_num$total_quantity, col = "pink", main = "Total Quantity")
```

## Log-Transformed Features

```{r}
agg_num <- agg_num %>%
  mutate(
    log_total_spend = log1p(total_spend),
    log_n_orders = log1p(n_orders),
    log_total_quantity = log1p(total_quantity),
    spend_per_order = total_spend / n_orders,
    qty_per_order = total_quantity / n_orders
  ) %>%
  mutate(across(everything(), ~replace(., is.infinite(.), 0))) %>%
  mutate(across(everything(), ~replace_na(., 0)))

# summary(agg_num)
```


## Household Item Features

```{r}
household_keywords <- c(
  "TOILET", "PAPER", "TOWEL", "NAPKIN", "CLEAN", "SOAP", "LAUNDRY", "GROCERY", "FOOD", "DRINK",
  "WATER"
)

purch <- purch %>%
  mutate(
    Category_str = str_to_upper(as.character(Category)),
    household_item = map_lgl(Category_str, ~any(str_detect(., household_keywords)))
  )

household_counts <- purch %>%
  group_by(Survey.ResponseID) %>%
  summarise(n_household_items = sum(household_item, na.rm = TRUE)) %>%
  ungroup()

agg_num <- agg_num %>%
  left_join(household_counts, by = "Survey.ResponseID") %>%
  mutate(n_household_items = replace_na(n_household_items, 0))

# cat("Household items feature added\n")
```


## Kid/Baby/Children Item Features

```{r}
kid_keywords <- c(
  "DIAPER", "DIAPERS", "WIPES", "WIPE", "BABY", "INFANT", "TODDLER", "TOY", "TOYS", "STROLLER",
  "BOTTLE", "PACIFIER", "CRIB", "KIDS", "CHILD", "CHILDREN"
)

purch <- purch %>%
  mutate(
    kid_item = map_lgl(Category_str, ~any(str_detect(., kid_keywords)))
  )

kid_counts <- purch %>%
  group_by(Survey.ResponseID) %>%
  summarise(n_kid_items = sum(kid_item, na.rm = TRUE)) %>%
  ungroup()

agg_num <- agg_num %>%
  left_join(kid_counts, by = "Survey.ResponseID") %>%
  mutate(n_kid_items = replace_na(n_kid_items, 0))

# cat("Kid items feature added\n")
# cat("Users with kid items:", sum(agg_num$n_kid_items > 0), "\n")
```


```{r}
data <- survey %>%
  left_join(agg_num, by = "Survey.ResponseID")

TARGET_COL <- "Q.amazon.use.hh.size.num"
ID_COL <- "Survey.ResponseID"

# cat("Combined dataset shape:", nrow(data), "x", ncol(data), "\n")
```


## Average Household Items Purchased by Amazon Usage
```{r}
amazon_summary <- data %>%
  group_by(Q.amazon.use.hh.size.num) %>%
  summarise(avg_household_items = mean(n_household_items, na.rm = TRUE)) %>%
  ungroup()

ggplot(amazon_summary, aes(x = Q.amazon.use.hh.size.num, y = avg_household_items)) +
  geom_col(fill = "blue") +
  labs(
    title = "Average Household Items Purchased by Amazon Usage",
    x = "Q.amazon.use.hh.size",
    y = "Average Number of Household Items"
  ) +
  theme_minimal()
```

## Average Kid Items by Household Size
```{r}
amazon_summary <- data %>%
  group_by(Q.amazon.use.hh.size.num) %>%
  summarise(n_kid_items = mean(n_kid_items, na.rm = TRUE)) %>%
  ungroup()

ggplot(amazon_summary, aes(x = Q.amazon.use.hh.size.num, y = n_kid_items)) +
  geom_col(fill = "pink") +
  labs(
    title = "Average Household Items Purchased by Amazon Usage",
    x = "Q.amazon.use.hh.size",
    y = "Average Number of Household Items"
  ) +
  theme_minimal()
```

# 5. Train/Test Split

```{r}
train_df <- data %>% filter(test == 0)
test_df <- data %>% filter(test == 1)

y <- train_df[[TARGET_COL]]

# dropping comments 
cols_to_drop <- c(
  "Q.amazon.use.hh.size.num",   
  "Q.amazon.use.hh.size",      
  "Q.amazon.use.howmany",      
  "test",
  "Survey.ResponseID"
)

X <- train_df %>% select(-all_of(cols_to_drop))
X_test <- test_df %>% select(-all_of(cols_to_drop))

cat("Train shape:", nrow(X), "x", ncol(X), "\n")
cat("Test shape:", nrow(X_test), "x", ncol(X_test), "\n\n")

cat("Target distribution:\n")
print(table(y))
```

## Prepare Data for Modeling

```{r}
# char to factor 
X <- X %>% mutate(across(where(is.character), as.factor))
X_test <- X_test %>% mutate(across(where(is.character), as.factor))

for (col in names(X)) {
  if (is.factor(X[[col]])) {
    all_levels <- unique(c(levels(X[[col]]), levels(X_test[[col]])))
    X[[col]] <- factor(X[[col]], levels = all_levels)
    X_test[[col]] <- factor(X_test[[col]], levels = all_levels)
  }
}

```

## Handle Missing Values

```{r}
# check for NA
cat("Missing values in X:\n")
colSums(is.na(X)) %>% 
  as.data.frame() %>%
  filter(. > 0) %>%
  arrange(desc(.))

# numeric columns replace with median
# factor colunms replace with mode or create "Missing" level

X <- X %>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm = TRUE)))) %>%
  mutate(across(where(is.factor), ~fct_na_value_to_level(., level = "Missing")))

# NAs in test 
X_test <- X_test %>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm = TRUE)))) %>%
  mutate(across(where(is.factor), ~fct_na_value_to_level(., level = "Missing")))

complete_rows <- !is.na(y)
X <- X[complete_rows, ]
y <- y[complete_rows]

```

# 6. Final Model Training

## 6.1 Baseline Model


```{r}

mean_hh_size <- mean(y, na.rm = TRUE)

# cat("Mean household size:", round(mean_hh_size, 4), "\n")

# baseline predictions 
baseline_train_preds <- rep(mean_hh_size, length(y))
baseline_rmse <- sqrt(mean((y - baseline_train_preds)^2))

# cat("Baseline RMSE:", round(baseline_rmse, 4), "\n")

baseline_test_preds <- rep(mean_hh_size, nrow(X_test))
baseline_test_preds <- round(baseline_test_preds)
baseline_test_preds <- pmax(baseline_test_preds, 1)

table(baseline_test_preds)
```

### Baseline Model Individual Predictions

```{r}
example_indices <- 1:3
example_predictions <- tibble(
  Example = paste("Person", 1:3),
  `Actual Size` = y[example_indices],
  `Predicted Size` = rep(mean_hh_size, 3),
  Error = y[example_indices] - mean_hh_size
) %>%
  mutate(across(where(is.numeric), round, 2))

# bar plot 
example_long <- example_predictions %>%
  select(Example, `Actual Size`, `Predicted Size`) %>%
  pivot_longer(cols = -Example, names_to = "Type", values_to = "Value")

baseline_examples_plot <- ggplot(example_long, aes(x = Example, y = Value, fill = Type)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = round(Value, 2)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Actual Size" = "blue", 
                                "Predicted Size" = "pink")) +
  labs(
    title = "Baseline Model Examples",
    x = NULL,
    y = "Household Size",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40"),
    legend.position = "top"
  )

baseline_examples_plot
```

## 6.2 Lasso Regression

### Prepare Data for Lasso

```{r}

X_lasso <- model.matrix(~ . - 1, data = X)
X_test_lasso <- model.matrix(~ . - 1, data = X_test)

# cat("Lasso training matrix:", nrow(X_lasso), "x", ncol(X_lasso), "\n")
# cat("Lasso test matrix:", nrow(X_test_lasso), "x", ncol(X_test_lasso), "\n")
# cat("Length of y:", length(y), "\n")
```

### Train Lasso Model with Cross-Validation

```{r}
set.seed(42)

# cross-validation
cv_lasso <- cv.glmnet(
  x = X_lasso,
  y = y,
  alpha = 1,        
  nfolds = 10,
  type.measure = "mse"
)

# plot(cv_lasso, main = "Lasso Cross-Validation")

# cat("\nOptimal lambda (lambda.min):", cv_lasso$lambda.min, "\n")
# cat("Lambda with 1-SE rule (lambda.1se):", cv_lasso$lambda.1se, "\n")

# train final model 
lasso_model <- glmnet(
  x = X_lasso,
  y = y,
  alpha = 1,
  lambda = cv_lasso$lambda.min
)

lasso_train_preds <- predict(lasso_model, newx = X_lasso, s = cv_lasso$lambda.min)[,1]
lasso_train_rmse <- sqrt(mean((y - lasso_train_preds)^2))

# cat("Lasso Training RMSE:", round(lasso_train_rmse, 4), "\n")
```

### Lasso Feature Importance

```{r}
# remove none zero
lasso_coef <- coef(lasso_model, s = cv_lasso$lambda.min)
lasso_coef_df <- data.frame(
  Feature = rownames(lasso_coef),
  Coefficient = as.vector(lasso_coef)
) %>%
  filter(Coefficient != 0, Feature != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

# cat("\nNumber of non-zero coefficients:", nrow(lasso_coef_df), "out of", ncol(X_lasso), "\n")

# plot
lasso_coef_top <- lasso_coef_df %>%
  head(15) %>%
  mutate(Feature = fct_reorder(Feature, abs(Coefficient)))

ggplot(lasso_coef_top, aes(x = Coefficient, y = Feature, fill = Coefficient > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "darkblue", "FALSE" = "darkred"),
                    labels = c("Negative", "Positive")) +
  labs(
    title = "Lasso: Top 15 Features by Coefficient Magnitude",
    x = "Coefficient Value",
    y = NULL,
    fill = "Effect"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )

knitr::kable(
  lasso_coef_df %>% head(20),
  digits = 4,
  caption = "Top 20 Lasso Coefficients by Magnitude"
)
```

### Lasso Predictions

```{r}
lasso_test_preds <- predict(lasso_model, newx = X_test_lasso, s = cv_lasso$lambda.min)[,1]
```

### Lasso Individual Predictions 

```{r}
intercept <- as.numeric(coef(lasso_model, s = cv_lasso$lambda.min)[1])

for (i in 1:2) {
  # print prediction
  # cat(sprintf("\nPerson %d: Predicted household size = %.2f\n",
  #             i, lasso_test_preds[i]))
  
  # top 3 
  tibble(
    Feature = lasso_coef_df$Feature,
    Coef = lasso_coef_df$Coefficient,
    Value = X_test_lasso[i, lasso_coef_df$Feature],
    Contrib = Coef * Value
  ) %>%
    arrange(desc(abs(Contrib))) %>%
    head(3) %>%
    print()
  
  # cat(sprintf(
  #  "Base (%.2f) + Contributions = %.2f, rounds to %.0f\n",
  #  intercept,
  #  intercept + sum(X_test_lasso[i, lasso_coef_df$Feature] * lasso_coef_df$Coefficient),
  #  round(lasso_test_preds[i])
  # ))
}
```

	
## 6.3 Random Forest Model

### Train Random Forest

```{r}

set.seed(123)

# search for the best mtry value
tuned <- tuneRF(
  x = X,
  y = y,
  ntreeTry = 300,          
  stepFactor = 1.5,      
  improve = 0.01,         
  trace = TRUE,
  plot = TRUE
)

best_mtry <- tuned[which.min(tuned[,"OOBError"]), "mtry"]
# cat("Best mtry:", best_mtry, "\n")

set.seed(123)

rf_model <- randomForest(
  x = X,
  y = y,
  mtry = best_mtry,
  ntree = 500,         
  importance = TRUE
)

print(rf_model)
varImpPlot(rf_model)
```

### Random Forest Feature Importance

## Feature Importance

```{r}
importance_df <- randomForest::importance(rf_model) %>%
  as.data.frame() %>%
  rownames_to_column("Feature") %>%
  arrange(desc(IncNodePurity)) %>%
  head(20)

knitr::kable(importance_df, digits = 2, caption = "Top 20 Most Important Features")
```

### Random Forest Training Performance

## Training Performance

```{r}
train_preds <- predict(rf_model, X)
train_rmse <- sqrt(mean((y - train_preds)^2))

# cat("Training RMSE:", round(train_rmse, 4), "\n")
```

### Random Forest Feature Importance Visualization

# Feature Importance Visualization

```{r}
importance_df <- randomForest::importance(rf_model) %>%
  as.data.frame() %>%
  rownames_to_column("Feature") %>%
  arrange(desc(IncNodePurity)) %>%
  head(15) %>%
  mutate(Feature = fct_reorder(Feature, IncNodePurity))

ggplot(importance_df, aes(x = IncNodePurity, y = Feature)) +
  geom_col(fill = "blue") +
  labs(
    title = "Random Forest Feature Importance (Top 15)",
    x = "Importance (IncNodePurity)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text = element_text(size = 10)
  )
```

### Random Forest Individual Predictions 
```{r}
for (i in 1:2) {
  pred <- predict(rf_model, X_test[i,])
  
  top_features <- importance_df$Feature[1:3]
  
  cat(sprintf("\nPerson %d Prediction Explanation:\n", i))
  
  cat("Predicted HH Size = ", round(pred, 2), " approx ", round(pred), "\n")

  for (f in top_features) {
    value <- X_test[i, f]
    importance <- importance_df$IncNodePurity[importance_df$Feature == f]
    cat(sprintf("Feature: %s, Value: %s, Importance: %.2f\n", f, value, importance))
  }
  
  cat("\n")
}

```

## 6.4 XGBoost Model

### Prepare Data for XGBoost

```{r}
# numeric matrix
X_xgb <- model.matrix(~. - 1, data = X)
X_xgb_test <- model.matrix(~. - 1, data = X_test)

# ensure test set has all columns from training
missing_cols <- setdiff(colnames(X_xgb), colnames(X_xgb_test))
if (length(missing_cols) > 0) {
  for (col in missing_cols) {
    X_xgb_test <- cbind(X_xgb_test, 0)
    colnames(X_xgb_test)[ncol(X_xgb_test)] <- col
  }
}
X_xgb_test <- X_xgb_test[, colnames(X_xgb)]

# dmatrix objects
dtrain <- xgb.DMatrix(data = X_xgb, label = y)
dtest <- xgb.DMatrix(data = X_xgb_test)

cat("XGBoost data prepared\n")
cat("Training matrix dimensions:", dim(X_xgb), "\n")
cat("Test matrix dimensions:", dim(X_xgb_test), "\n")
```

### Hyperparameter Tuning for XGBoost

```{r}
set.seed(123)

# hyperparameter grid
param_grid <- expand.grid(
  max_depth = c(3, 5, 7),
  eta = c(0.01, 0.1, 0.3),
  subsample = c(0.7, 0.8, 1.0),
  colsample_bytree = c(0.7, 0.8, 1.0)
)

# cat("Testing", nrow(param_grid), "hyperparameter combinations\n\n")

tuning_results <- tibble()

# rest 
for (i in 1:min(10, nrow(param_grid))) {  # Test first 10 combinations for efficiency
  params <- list(
    objective = "reg:squarederror",
    eval_metric = "rmse",
    max_depth = param_grid$max_depth[i],
    eta = param_grid$eta[i],
    subsample = param_grid$subsample[i],
    colsample_bytree = param_grid$colsample_bytree[i]
  )
  
  cv_result <- xgb.cv(
    params = params,
    data = dtrain,
    nrounds = 100,
    nfold = 5,
    early_stopping_rounds = 10,
    verbose = FALSE
  )
  
  best_rmse <- min(cv_result$evaluation_log$test_rmse_mean)
  
  tuning_results <- bind_rows(tuning_results, tibble(
    max_depth = param_grid$max_depth[i],
    eta = param_grid$eta[i],
    subsample = param_grid$subsample[i],
    colsample_bytree = param_grid$colsample_bytree[i],
    cv_rmse = best_rmse
  ))
  
}

# select best parameters
best_params_row <- tuning_results %>% arrange(cv_rmse) %>% slice(1)

cat("\nBest Hyperparameters:\n")
print(best_params_row)
```

### Train Final XGBoost Model with Best Parameters

```{r}
set.seed(123)

best_params <- list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  max_depth = best_params_row$max_depth,
  eta = best_params_row$eta,
  subsample = best_params_row$subsample,
  colsample_bytree = best_params_row$colsample_bytree
)

xgb_model <- xgb.train(
  params = best_params,
  data = dtrain,
  nrounds = 200,
  verbose = FALSE
)

xgb_train_preds <- predict(xgb_model, dtrain)
xgb_train_rmse <- sqrt(mean((y - xgb_train_preds)^2))

# cat("XGBoost Model Training Complete\n")
# cat("Training RMSE:", round(xgb_train_rmse, 4), "\n")
```

### XGBoost Feature Importance

```{r}
xgb_importance <- xgb.importance(model = xgb_model)

cat("\nTop 10 Most Important Features (XGBoost):\n")
print(head(xgb_importance, 10))

xgb.plot.importance(head(xgb_importance, 15), 
                    main = "Top 15 Features by Importance (XGBoost)")
```

### XGBoost Individual Predictions

```{r}
for (i in 1:2) {
  pred <- predict(xgb_model, X_xgb_test[i, , drop = FALSE])
  
  top_features <- xgb_importance$Feature[1:3]
  
  cat(sprintf("\nPerson %d Prediction Explanation:\n", i))
  cat("Predicted HH Size =", round(pred, 2), " ", round(pred), "\n")
  
  for (f in top_features) {
    if (f %in% colnames(X_xgb_test)) {
      value <- X_xgb_test[i, f]
      gain <- xgb_importance$Gain[xgb_importance$Feature == f]
      cat(sprintf("Feature: %s, Value: %.2f, Gain: %.4f\n", f, value, gain))
    }
  }
  cat("\n")
}
```

# 7. K-Fold Cross-Validation Setup

```{r}
set.seed(123)

k_folds <- 5

n <- nrow(X)
fold_indices <- sample(rep(1:k_folds, length.out = n))

X_lasso_full <- model.matrix(~ . - 1, data = X)

cat("Fold sizes:\n")
print(table(fold_indices))
```

## 7.1 Cross-Validation Function

```{r}
calculate_rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

# kfold CV for baseline 
cv_baseline <- function(X, y, fold_indices, k_folds) {
  rmse_scores <- numeric(k_folds)
  
  for (fold in 1:k_folds) {
    train_idx <- fold_indices != fold
    val_idx <- fold_indices == fold
    
    y_train <- y[train_idx]
    y_val <- y[val_idx]
    
    mean_train <- mean(y_train)
    val_preds <- rep(mean_train, sum(val_idx))
    
    rmse_scores[fold] <- calculate_rmse(y_val, val_preds)
  }
  
  return(rmse_scores)
}

# kfold CV for lasso
cv_lasso_kfold <- function(X_lasso, y, fold_indices, k_folds) {
  rmse_scores <- numeric(k_folds)
  
  for (fold in 1:k_folds) {
    train_idx <- fold_indices != fold
    val_idx <- fold_indices == fold
    
    X_train <- X_lasso[train_idx, ]
    X_val <- X_lasso[val_idx, ]
    y_train <- y[train_idx]
    y_val <- y[val_idx]
    
    cv_model <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 5)
    
    val_preds <- predict(cv_model, newx = X_val, s = "lambda.min")[,1]
    
    rmse_scores[fold] <- calculate_rmse(y_val, val_preds)
  }
  
  return(rmse_scores)
}

# kfold CV for random forest
cv_random_forest <- function(X, y, fold_indices, k_folds, mtry = NULL, ntree = 300) {
  rmse_scores <- numeric(k_folds)
  
  if (is.null(mtry)) {
    mtry <- floor(sqrt(ncol(X)))
  }
  
  for (fold in 1:k_folds) {
    train_idx <- fold_indices != fold
    val_idx <- fold_indices == fold
    
    X_train <- X[train_idx, ]
    X_val <- X[val_idx, ]
    y_train <- y[train_idx]
    y_val <- y[val_idx]
    
    rf_model <- randomForest(
      x = X_train,
      y = y_train,
      mtry = mtry,
      ntree = ntree,
      importance = FALSE
    )
    
    val_preds <- predict(rf_model, X_val)
    rmse_scores[fold] <- calculate_rmse(y_val, val_preds)
    cat("Fold", fold, "- RMSE:", round(rmse_scores[fold], 4), "\n")
  }
  
  return(rmse_scores)
}

# kfold CV for xgboost
cv_xgboost <- function(X_xgb, y, fold_indices, k_folds, params) {
  rmse_scores <- numeric(k_folds)
  
  for (fold in 1:k_folds) {
    train_idx <- fold_indices != fold
    val_idx <- fold_indices == fold
    
    X_train <- X_xgb[train_idx, ]
    X_val <- X_xgb[val_idx, ]
    y_train <- y[train_idx]
    y_val <- y[val_idx]
    
    dtrain_fold <- xgb.DMatrix(data = X_train, label = y_train)
    dval_fold <- xgb.DMatrix(data = X_val, label = y_val)
    
    xgb_fold <- xgb.train(
      params = params,
      data = dtrain_fold,
      nrounds = 200,
      verbose = FALSE
    )
    
    val_preds <- predict(xgb_fold, dval_fold)
    
    rmse_scores[fold] <- calculate_rmse(y_val, val_preds)
    
    cat("Fold", fold, "- RMSE:", round(rmse_scores[fold], 4), "\n")
  }
  return(rmse_scores)
}
```

## 7.2 Perform K-Fold Cross-Validation for All Models

### Baseline Model CV

```{r}
baseline_cv_scores <- cv_baseline(X, y, fold_indices, k_folds)

cat("Fold RMSEs:", paste(round(baseline_cv_scores, 4), collapse = ", "), "\n")
cat("Mean CV RMSE:", round(mean(baseline_cv_scores), 4), "\n")
cat("Std CV RMSE:", round(sd(baseline_cv_scores), 4), "\n")
```

### Lasso Model CV

```{r}
lasso_cv_scores <- cv_lasso_kfold(X_lasso_full, y, fold_indices, k_folds)

cat("Fold RMSEs:", paste(round(lasso_cv_scores, 4), collapse = ", "), "\n")
cat("Mean CV RMSE:", round(mean(lasso_cv_scores), 4), "\n")
cat("Std CV RMSE:", round(sd(lasso_cv_scores), 4), "\n")
```

### Random Forest Model CV

```{r}
set.seed(123)
tuned_sample <- tuneRF(
  x = X,
  y = y,
  ntreeTry = 300,
  stepFactor = 1.5,
  improve = 0.01,
  trace = FALSE,
  plot = FALSE
)
optimal_mtry <- tuned_sample[which.min(tuned_sample[,"OOBError"]), "mtry"]

cat("Using mtry =", optimal_mtry, "for Random Forest CV\n\n")

# kfold 
rf_cv_scores <- cv_random_forest(X, y, fold_indices, k_folds, 
                                  mtry = optimal_mtry, ntree = 300)

cat("Fold RMSEs:", paste(round(rf_cv_scores, 4), collapse = ", "), "\n")
cat("Mean CV RMSE:", round(mean(rf_cv_scores), 4), "\n")
cat("Std CV RMSE:", round(sd(rf_cv_scores), 4), "\n")
```

### XGBoost Model CV

```{r}
cat("Using best hyperparameters from tuning\n\n")

xgb_cv_scores <- cv_xgboost(X_xgb, y, fold_indices, k_folds, best_params)

cat("\nFold RMSEs:", paste(round(xgb_cv_scores, 4), collapse = ", "), "\n")
cat("Mean CV RMSE:", round(mean(xgb_cv_scores), 4), "\n")
cat("Std CV RMSE:", round(sd(xgb_cv_scores), 4), "\n")
```

## 7.3 Cross-Validation Results Summary

```{r}
# comparison dataframe
cv_results <- tibble(
  Model = c("Baseline (Mean)", "Lasso", "Random Forest", "XGBoost"),
  `Mean CV RMSE` = c(
    mean(baseline_cv_scores),
    mean(lasso_cv_scores),
    mean(rf_cv_scores),
    mean(xgb_cv_scores)
  ),
  `Std CV RMSE` = c(
    sd(baseline_cv_scores),
    sd(lasso_cv_scores),
    sd(rf_cv_scores),
    sd(xgb_cv_scores)
  ),
  `Min CV RMSE` = c(
    min(baseline_cv_scores),
    min(lasso_cv_scores),
    min(rf_cv_scores),
    min(xgb_cv_scores)
  ),
  `Max CV RMSE` = c(
    max(baseline_cv_scores),
    max(lasso_cv_scores),
    max(rf_cv_scores),
    max(xgb_cv_scores)
  )
)

knitr::kable(
  cv_results,
  digits = 4,
  caption = "5-Fold Cross-Validation Results",
  align = "lrrrr"
)

# rank models by CV 
cv_results %>%
  arrange(`Mean CV RMSE`) %>%
  select(Model, `Mean CV RMSE`) %>%
  knitr::kable(digits = 4)
```

### Visualization of CV Results

```{r}

cv_plot_data <- tibble(
  Model = rep(c("Baseline", "Lasso", "Random Forest", "XGBoost"), each = k_folds),
  Fold = rep(1:k_folds, 4),
  RMSE = c(baseline_cv_scores, lasso_cv_scores, rf_cv_scores, xgb_cv_scores)
)

# boxplot of CV scores
ggplot(cv_plot_data, aes(x = Model, y = RMSE, fill = Model)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  labs(
    title = "Cross-Validation RMSE by Model",
    subtitle = paste0(k_folds, "-Fold Cross-Validation Results"),
    x = "Model",
    y = "RMSE"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    legend.position = "none"
  ) +
  scale_fill_brewer(palette = "Set2")

# line plot
ggplot(cv_plot_data, aes(x = Fold, y = RMSE, color = Model, group = Model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Model Performance Across CV Folds",
    x = "Fold Number",
    y = "RMSE"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_color_brewer(palette = "Set1")
```

# 8. Model Comparison Summary

```{r}
# compare with tables 
model_comparison <- tibble(
  Model = c("Baseline (Mean)", "Lasso", "Random Forest", "XGBoost"),
  `Training RMSE` = c(
    round(baseline_rmse, 4),
    round(lasso_train_rmse, 4),
    round(train_rmse, 4),
    round(xgb_train_rmse, 4)
  ),
  `Mean CV RMSE` = c(
    round(mean(baseline_cv_scores), 4),
    round(mean(lasso_cv_scores), 4),
    round(mean(rf_cv_scores), 4),
    round(mean(xgb_cv_scores), 4)
  ),
  `Std CV RMSE` = c(
    round(sd(baseline_cv_scores), 4),
    round(sd(lasso_cv_scores), 4),
    round(sd(rf_cv_scores), 4),
    round(sd(xgb_cv_scores), 4)
  ),
  `Number of Features` = c(
    1,  
    nrow(lasso_coef_df), 
    ncol(X),
    ncol(X_xgb)
  ),
  Notes = c(
    "Predicts mean for all observations",
    "Linear model with L1 regularization",
    "Ensemble of 500 decision trees",
    "Gradient boosted trees with tuned hyperparameters"
  )
)

knitr::kable(
  model_comparison,
  caption = "Model Performance Comparison (Training vs Cross-Validation)",
  align = c("l", "r", "r", "r", "r", "l")
)

cat("Model Rankings by Cross-Validation RMSE\n")
model_comparison %>%
  arrange(`Mean CV RMSE`) %>%
  select(Model, `Mean CV RMSE`, `Std CV RMSE`) %>%
  knitr::kable(digits = 4)


cat("\nImprovement Over Baseline\n")
baseline_cv_mean <- mean(baseline_cv_scores)
lasso_cv_mean <- mean(lasso_cv_scores)
rf_cv_mean <- mean(rf_cv_scores)
xgb_cv_mean <- mean(xgb_cv_scores)

cat("Lasso improvement:", round((1 - lasso_cv_mean/baseline_cv_mean) * 100, 2), "%\n")
cat("Random Forest improvement:", round((1 - rf_cv_mean/baseline_cv_mean) * 100, 2), "%\n")
cat("XGBoost improvement:", round((1 - xgb_cv_mean/baseline_cv_mean) * 100, 2), "%\n")
```

# 9. Generate Final Predictions (Best Model)

```{r}
# best model 
best_model_name <- model_comparison %>%
  arrange(`Mean CV RMSE`) %>%
  slice(1) %>%
  pull(Model)

cat("Best Model by Cross-Validation:", best_model_name, "\n\n")

# generate predictions with best performing model
if (best_model_name == "XGBoost") {
  test_preds <- predict(xgb_model, dtest)
} else if (best_model_name == "Random Forest") {
  test_preds <- predict(rf_model, X_test)
} else if (best_model_name == "Lasso") {
  test_preds <- predict(lasso_model, newx = X_lasso_test_scaled)[,1]
} else {
  test_preds <- rep(mean_y, nrow(X_test))
}

# household size must be at least 1
test_preds <- pmax(test_preds, 1)

kaggle <- tibble(
  Survey.ResponseID = test_df$Survey.ResponseID,
  Q.amazon.use.hh.size.num = test_preds
)

write_csv(kaggle, "submission.csv")
cat("Submission file saved as 'submission.csv'\n")
cat("Using predictions from:", best_model_name, "\n\n")
cat("Prediction summary:\n")
head(table(round(test_preds)))
head(kaggle)
```

# 10. Summary Statistics

```{r}
cat("BASELINE MODEL:\n")
cat("  Training RMSE:", round(baseline_rmse, 4), "\n")
cat("  CV RMSE:", round(mean(baseline_cv_scores), 4), "\n")
cat("  Method: Mean prediction\n\n")

cat("LASSO REGRESSION:\n")
cat("  Training RMSE:", round(lasso_train_rmse, 4), "\n")
cat("  CV RMSE:", round(mean(lasso_cv_scores), 4), "\n")
cat("  Optimal lambda:", round(cv_lasso$lambda.min, 6), "\n")
cat("  Non-zero coefficients:", nrow(lasso_coef_df), "out of", ncol(X_lasso), "\n\n")

cat("RANDOM FOREST MODEL:\n")
cat("  Training RMSE:", round(train_rmse, 4), "\n")
cat("  CV RMSE:", round(mean(rf_cv_scores), 4), "\n")
cat("  Number of trees:", rf_model$ntree, "\n")
cat("  Number of features:", ncol(X), "\n\n")

cat("XGBOOST MODEL:\n")
cat("  Training RMSE:", round(xgb_train_rmse, 4), "\n")
cat("  CV RMSE:", round(mean(xgb_cv_scores), 4), "\n")
cat("  Best hyperparameters:\n")
cat("    max_depth:", best_params_row$max_depth, "\n")
cat("    eta:", best_params_row$eta, "\n")
cat("    subsample:", best_params_row$subsample, "\n")
cat("    colsample_bytree:", best_params_row$colsample_bytree, "\n\n")

cat("RMSE Improvement over Baseline (CV):\n")
cat("  Lasso:", round((1 - lasso_cv_mean/baseline_cv_mean) * 100, 2), "%\n")
cat("  Random Forest:", round((1 - rf_cv_mean/baseline_cv_mean) * 100, 2), "%\n")
cat("  XGBoost:", round((1 - xgb_cv_mean/baseline_cv_mean) * 100, 2), "%\n\n")

cat("BEST MODEL:", best_model_name, "\n")
cat("  Mean CV RMSE:", round(min(model_comparison$`Mean CV RMSE`), 4), "\n")
```